---
title: "Functional analysis of the Senescence project (builds upon results in Bulk DE analysis)"
date: "`r Sys.Date()`"
author: "Anna Lovrics"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```


```{r libraries}

# replace with new packages!!!
# install CRAN packages via BioConductor as well
if (!"BiocManager" %in% installed.packages()[,"Package"]){
  install.packages("BiocManager")
}
list.of.packages <- c("ggplot2", "limma", "dplyr", "tidyr", "openxlsx", 
                      "fgsea", "msigdbr", "stringr")
new.packages <- setdiff(list.of.packages, installed.packages()[,"Package"])
if(length(new.packages)) BiocManager::install(new.packages)

library(ggplot2) # for nice plots
library(limma) # for the diff expression analysis
library(dplyr) # to work with tibbles & other data management
library(tidyr) # for data manipulation
library(openxlsx) # to read and write data from/to xlsx files
library(fgsea)
library(msigdbr) # for the gene sets
library(stringr) # for str_wrap
```

```{r folder structure}
# set the  working directory the directory you launched R from
setwd(Sys.getenv("PWD")) 
mainfolder <- getwd()
datafolder <- file.path(mainfolder, "Data")
calcfolder <- file.path(mainfolder, "Data_calculated")
figfolder <- file.path(mainfolder, "Figures")
resfolder <- file.path(mainfolder, "Results")
```


```{r read in normalised expression and gene annotation}
# read in differential expression result
outfile <- file.path(resfolder, "limma_fit_diffexpr.rds")
fite <- readRDS(file=outfile)

infile <- file.path(datafolder, "Samples/gene_annotation.csv")
genes_annot <- read.delim(infile,sep="\t")[,c('Gene_ID', 'name', 'description')]
genes_annot$label <- ifelse(genes_annot$name == "None", 
                            genes_annot$Gene_ID,
                            genes_annot$name)
```

Select contrasts: TIS vs CTR, REPOP vs TIS in the malignant cell lines and TIS vs CTR in HFF cells
```{r create list of ranks using logFC (instead of p.adjust as before)}
mcf_vector <- c("control2senescent_mal", "senescent2repopulated_mal",
                 "control2senescent_HFF")
toptt_orig <- list()
toptt_ranks <- list()
#postfix <- "padj"
postfix <- "logFC"
for (mcf in mcf_vector){
  toptt_orig[[mcf]] <- topTable(fite, coef=mcf,  
                    number=nrow(fite$coefficients), sort.by = "none")
  if (postfix == "logFC"){
    toptt_ranks[[mcf]] <- toptt_orig[[mcf]]$logFC
    toptt_ranks[[mcf]] <- as.vector(scale(toptt_ranks[[mcf]], center=T, scale=F))
  }
 if (postfix == "padj"){
    toptt_ranks[[mcf]] <- -log10(toptt_orig[[mcf]]$adj.P.Val)*sign(toptt_orig[[mcf]]$logFC)
  }
  names(toptt_ranks[[mcf]]) <- rownames(toptt_orig[[mcf]])
}
```

GSEA analysis for the contrasts with selected gene sets
```{r GSEA analysis}
# selected gene sets (GO:BP, CP:KEGG, hallmark)
category_names <- c("C5", "C2", "H")
subcategory_names <- c("GO:BP","CP:KEGG", "")
prefix <- c("gobp_", "kegg_", "hallmark_")
# repeat for different coefficients
FA_lists <- list()
for (mcf in mcf_vector){
  folderFA <-  file.path(resfolder, 
                 paste("FA", mcf, postfix, sep="_"))
  if (!file.exists(folderFA)) dir.create(folderFA)
  geneset_file <- file.path(folderFA, "GO_category_enrichments.rds")
  if (file.exists(geneset_file)){
    FA_lists[[mcf]] <- readRDS(geneset_file)
  } else {
  # repeat for different gene sets
  for (gsi in c(1:length(category_names))){
  #for (gsi in c(1:1)){
    broad_name <- subcategory_names[gsi]
    folder_name <- ifelse(nchar(broad_name) > 0, broad_name, category_names[gsi])
    # create a folder
    folderenrres <- file.path(folderFA, folder_name)
    if (!file.exists(folderenrres)) dir.create(folderenrres)
    # use the actual genesets
    GO_db <- msigdbr(species= "Homo sapiens",
                     category=category_names[gsi], subcategory = subcategory_names[gsi])
    GO_db_list <- unique(GO_db[, c("ensembl_gene","gs_name")])
    GO_db_list <- split(x=GO_db_list$ensembl_gene, f=GO_db_list$gs_name)
    print(paste("GO_db_list ", folder_name, " created", sep = ""))
    # remove genes not in universe
    for (gsname in names(GO_db_list)){
      # keepgene <- which(GO_db_list[[gsname]] %in% names(fite$Amean))
      keepgene <- which(GO_db_list[[gsname]] %in% rownames(toptt_orig[[mcf]]))
      GO_db_list[[gsname]] <- GO_db_list[[gsname]][keepgene]
    }
    # now the gene enrichment analysis
    # genes_universe <- names(fite$Amean)
    genes_universe <- rownames(toptt_orig[[mcf]])
    set.seed(99)
    system.time( # keep track of elapsed time
      fgsea_res <- fgseaMultilevel(pathways = GO_db_list,
                                   stats = toptt_ranks[[mcf]], 
                                   minSize = 0, 
                                   maxSize = 500, 
                                   eps = 0, 
                                   nPermSimple = 50000)
    )
    fgsea_res <- arrange(fgsea_res, -NES)
    # remove na rows
    omit <- which(is.na(fgsea_res$pval))
    if (length(omit) > 0) fgsea_res <- fgsea_res[-omit,]
    # write out leading edge genes
    fgsea_res$leadingSymbols <- fgsea_res$leadingEdge
    for (ri in 1:nrow(fgsea_res)){
      gsymbols <- genes_annot[match(fgsea_res$leadingEdge[[ri]], genes_annot$Gene_ID),"label"]
      fgsea_res$leadingEdge[[ri]] <- gsymbols
    }
    
    # add number of leading edge genes
    fgsea_res$count = lengths(fgsea_res$leadingEdge)
    
    # add gene ratio & background ratio 
    Nval <- nrow(fite$coefficients)
    nval <- length(which(toptt_orig[[mcf]]$adj.P.Val<=0.05 & 
                           toptt_orig[[mcf]]$logFC>=1))
    fgsea_res$GeneRatio <- fgsea_res$count/nval
    fgsea_res$BgRatio <- fgsea_res$size/Nval
    
    # write to file
    outfile <- file.path(folderenrres, 
                        paste("enriched_genesets_fgsea_", mcf, ".xlsx", sep=""))
    write.xlsx(fgsea_res, file = outfile, colNames=T, rowNames=F)
    
    # put into the list
    FA_lists[[mcf]][[folder_name]] <- fgsea_res
  } # repeated for different gene sets
    saveRDS(FA_lists[[mcf]], file=geneset_file)
  } # end for check whether geneset_file exists
} # end of loop for different contrasts (ctr2sen_mal, sen2repop_mal, ctr2sen_HFF)

```

# Visual representation of gene set enrichment results of "H", GO:BP and KEGG pathways. Included graphics for Figure 4 L, M (dotplots for contrasts TIS vs CTR and REPOP vs TIS in malignant cells for hallmark gene sets), Supplementary Figure S3E (dotplot for contrast TIS vs CTR in HFF for hallmark gene sets), Figure S3F (dotplot for contrast TIS vs CTR in malignant cells for KEGG gene sets), Figure S3G (dotplot for contrast TIS vs CTR in HFF for KEGG gene sets)
```{r dotplot for selected gene sets}
for (mcf in mcf_vector){
  geneset_file <- file.path(folderFA, "GO_category_enrichments.rds")
  for (gsi in c(1:length(category_names))){
    y <- FA_lists[[mcf]][[gsi]]
    
    # select activated and supressed genesets
    y$direction <- "activated"
    y$direction[y$NES<=0] <- "suppressed"
    numact <- length(which(y$direction=="activated"))
    numsupp <- length(which(y$direction=="suppressed"))

    # pick top and bottom genesets
    act_Nvals <- y$NES[y$direction=="activated"]
    act_thr <- sort(act_Nvals, decreasing = T)[min(10, numact)]
    act_keep <- which(y$direction=="activated" & y$NES >= act_thr)
    sup_Nvals <- y$NES[y$direction=="suppressed"]
    sup_thr <- sort(sup_Nvals, decreasing = F)[min(10, numsupp)]
    sup_keep <- which(y$direction=="suppressed" & y$NES <= sup_thr)
    y <- y[c(act_keep, sup_keep),]
    
    # sort by normalized enrichment
    y <- y[order(y$NES, decreasing = F),]
    # change the names of the gene sets
    y$pathway <- tolower(y$pathway)
    y$pathway <- gsub(prefix[gsi], "", y$pathway)
    y$pathway <- gsub("_", " ", y$pathway)
    y$pathway <- str_to_sentence(y$pathway)
    y$pathway <- gsub("dna", "DNA", y$pathway)
    y$pathway <- gsub("Dna", "DNA", y$pathway)
    #y$pathway <- nice_names$name[match(y$pathway, nice_names$compare)]
    
    y$pathway <- factor(y$pathway, levels=y$pathway)
    
     # plot together
      p <- ggplot(y, 
                    aes(x = NES, y = pathway)) + 
               geom_point(aes(size = count, color = padj)) +
               guides(size=guide_legend(title="Count", ncol=1,byrow=TRUE,
                                title.vjust=0.9, order=1))+
               theme_bw(base_size = 12) +
               scale_color_continuous(low="red", high="blue", name = "p.adjust",
                     guide=guide_colorbar(reverse=TRUE))+
               theme(legend.position="right", 
                      legend.box = "vertical",
                     #axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
                     #strip.background =element_rect(fill=cfill),
                     strip.text = element_text(colour = 'black'))+
               ylab(NULL) + xlab("NES")+ 
               scale_size(range=c(3, 8))+
               #scale_y_discrete(labels = function(x) str_wrap(x, width = 24))+
               geom_vline(xintercept = 0, linetype="dashed")
    figfile <- file.path(figfolder, paste(mcf, '_',
           names(FA_lists[[mcf]])[gsi], '_', postfix, '_', 'both',
           "_enrich_dotplot.pdf", sep=""))
    ggsave(figfile, plot=p, width=7, height=6)

  }
}

```
